Symbolic Weighted Language Models and\\ Quantitative Parsing over Infinite Alphabets

### 2021-07-08 abstract.tex
Intro Philippe

% !TEX root = main.tex
%
% Symbolic Weighted (SW) extension of symbolic automata...
%
We propose a framework for weighted parsing over infinite alphabets.
%
It is based on language models called Symbolic Weighted Automata (SWA) 
at the joint %intersection 
between Symbolic Automata (SA) and Weighted Automata (WA), 
as well as Transducers (SWT) and Visibly Pushdown (SWPDA) variants.
%
Like SA, SWA deal with large or infinite input alphabets, 
and like WA, they output a weight value in a semiring domain.
The transitions of WA are labeled by functions from an infinite alphabet into the weight domain. 
This is unlike WA whose transitions are guarded by boolean predicates 
overs symbols in an infinite alphabet 
and also unlike WA whose transitions are labeled by constant weight values, 
and who deal only with finite automata.
%
We present some properties of WA, WT and SWPDA models, 
%and show how they can be used 
that we use to define and solve a variant of parsing 
over infinite alphabets.
%
We also briefly describe the application that motivated the introduction of these models: 
a parse-based approach to automated music transcription.

>>>>

% !TEX root = main.tex
%
% Symbolic Weighted (SW) extension of symbolic automata...
%
We propose a framework for weighted parsing over infinite alphabets.
%
It is based on language models called Symbolic Weighted Automata (\SWA)
at the joint %intersection
between Symbolic Automata (\SA) and Weighted Automata (\WA),
as well as Transducers (\SWT) and Visibly Pushdown (\SWVPA) variants.
%
Like \SA, \SWA deal with large or infinite input alphabets,
and like \WA, they output a weight value in a semiring domain.
The transitions of \SWA are labeled by functions from an infinite alphabet into the weight domain.
This is unlike \SA whose transitions are guarded by boolean predicates
overs symbols in an infinite alphabet
and also unlike \WA whose transitions are labeled by constant weight values,
and who deal only with finite automata.
%
We present some properties of \SWA, \SWT and \SWVPA models,
%and show how they can be used
that we use to define and solve a variant of parsing
over infinite alphabets.
%
We illustrate the models with examples taken from
a motivating application, namely
a parse-based approach to automated music transcription.

### 2021-07-08 introduction.tex
Intro Philippe

% !TEX root = main.tex
%\todo
% Introduction
%
Parsing %variant of membership problem for formal languages
%- given a language model and an input word $s$, compute a derivation of the model that yields $s$.
is the problem %process 
of structuring a linear representation on input %typically
(a finite word), according to a language model. % (a formal grammar).
% natural language, programming language, 
%
Most of the context-free parsing approaches~\cite{GruneJacobs08parsing}  
assume a finite and reasonably small input alphabet. %models and algorithms 
Such a restriction makes perfect sense in the context of NLP tasks such as constituency parsing,
or of programming languages compilers or interpreters.
Considering large or infinite alphabets can however be of practical interest, %in other cases.
%
for instance, when dealing with large characters encodings such as UTF-16, % processing strings in 
\eg for vulnerability detection in Web-applications~\cite{dAntoni21CACM}, 
%
for the analyse (\eg validation or filtering) 
of data streams or serialization of structured documents 
(with textual or numerical attributes)~\cite{Segoufin06csl}, 
or for processing timed execution traces~\cite{Bouyer03algebraic}.
% algebraic definition of a class of data languages 
% (notion of monoid recognizability, based on registers, comparable to Bojancszik et al. data words)
%
The latter case is related to a study that motivated the present work: 
automated music transcription.
\philippe{I think that we can make the case for a larger set of situations: given a linear 
music notation input, for instance elements in an XML file, we we want to structure
the input according to a hierarchical rhythmic space. I can elaborate...}
In this problem, a music performance, % a symbolic representation of a music performance 
represented symbolically in the form of a sequence of timed musical events, % timestamped
is converted into a score in Common Western Music Notation~\cite{Gould11Notation},
structured according to nested grouping and metric strength of events.
%(the structure represents hierarchical  grouping of events, and their relative temporal positions)
It can therefore be stated as a parsing problem~\cite{foscarin:hal-01988990}, 
over an infinite alphabet of timed events.

@@

Other extensions of language models  %(automata and grammars) 
help in dealing with non-determinism, by the computation of weight values. 
With an ambiguous grammar, there may exist several derivations 
(\emph{abstract syntax trees} -- AST) % \ie the result of an analyze.
yielding one input word. % structuring a word, 
The association of one weight value %in semiring domain
to each AST permits to select a best one (or $n$ bests). % a fixed number of bests 
% = ...a ranking of ASTs 
This is roughly the principle of \emph{weighted parsing}
approaches~\cite{Goodman99SemiringParsing,Nederhof03weightedParsing,MorbitzVogler19weighted-parsing}.
In \emph{weighted language models}, 
like \eg probabilistic context-free grammars % (CFG),
and weighted automata (\WA) \cite{Droste09handbook},
a weight value is associated to each transition rule, % production rule
and the rule's weights can be combined with a associative product operator~$\otimes$ into 
the weight of an AST.
A second operator~$\oplus$, associative and commutative, 
is moreover used to handle the ambiguity of the model, 
by summing the weights of the possibly several (in general exponentially many) AST %syntax trees 
associated to a given input word.
Typically, $\oplus$ will select the best of two weight values.
%Intuitively,~$\oplus$ selects, or ranking, the syntax trees.
The weight domain, equipped with these two operators shall be, at minima, 
a \emph{semiring} %$\Semiring$ 
where $\oplus$ can be extended to infinite sums, 
%\cite{Eilenberg74automata}
such as the Viterbi semiring and the tropical min-plus algebra, see Figure~\ref{fig:semirings}. 
%of domain $\mathbb{R}^+ \cup \{ +\infty\}$, 
%where $\oplus$ is min and $\otimes$ is plus .
%by ranking 
%making the weight domain a semiring.
%Some efficient specialized parsing algorithms~\cite{Huang05kbest} have been proposed in this context 
%% models represented as hypergraphs \cite{Huang05kbest}
%in order to compute the $n$ best syntax trees of a given input word without having to enumerate them all.
%Generally based on dynamic programming, these algorithms rely on 
%additional algebraic properties of~$\Semiring$.
%-- see \eg~\cite{Huang05kbest} for some NLP applications.
%The extraction of $n$ best list is useful 
%the problem: quantitative parsing or symbolic parsing
%of parsing of words over infinite input alphabet.

In this paper, we present a uniform framework for weighted parsing over infinite input alphabets.
It is based on \emph{symbolic weighted} finite states language models (\swM), 
generalizing~\SA, with functions into an arbitrary semiring instead of Boolean guards, 
and~\WA, by handling infinite alphabets, see Figure~\ref{fig:hierarchy}.
%***
%some weighted language models
%model of weighted CFG 
%computing on words over infinite input alphabets %sets of terminal symbol
%- but with finite sets of states and transitions rules. 
% non-terminals and production rules).
In the transition rules of \swM models, input symbols appear as variables, %(parameters)
and the weight associated to a transition rule is a function of these variables.
%
\begin{figure}
\centering
\begin{tikzpicture}
\node (NFA) at (0,4) {%
  \(
  \begin{array}{c} 
  \mathsf{FA} : \Sigma_{\mathsf{fin}}^* \to \mathbb{B}\\
  q \xrightarrow{a} q' \quad  a \in \Sigma_\mathsf{fin}
  \end{array} 
  \)
};
%
\node (WA) at (-2.5,2) {%
  \(
  \begin{array}{c} 
  \mathsf{WA} : \Sigma_{\mathsf{fin}}^* \to \mathbb{S}\\
  q \xrightarrow{a, w} q' \quad  a \in \Sigma_\mathsf{fin}, w \in \mathbb{S}
  \end{array} 
  \)
};
%
\node (SA) at (2.5,2) {%
  \(
  \begin{array}{c} 
  \mathsf{SA} : \Sigma_{\mathsf{inf}}^* \to \mathbb{B}\\
  q \xrightarrow{\phi} q' \quad \phi : \Sigma_\mathsf{inf} \to \mathbb{B}
  \end{array} 
  \)
};
%
\node (SWA) at (0,0) {%
  \(
  \begin{array}{c} 
  \mathsf{SWA} : \Sigma_{\mathsf{inf}}^* \to \mathbb{S}\\
  q \xrightarrow{\phi} q' \quad \phi : \Sigma_\mathsf{inf} \to \mathbb{S}
  \end{array} 
  \)
};
\draw[->] (NFA)--(WA);
\draw[->] (NFA)--(SA);
\draw[->] (WA)--(SWA);
\draw[->] (SA)--(SWA);
%\begin{array}{c} \mathsf{NFA} : \Sigma^* \to \mathbb{B} \end{array} 
\end{tikzpicture}
\caption{Classes of Symbolic/Weighted Automata. 
$\Sigma_\mathsf{fin}$ is a finite alphabet, 
$\Sigma_\mathsf{inf}$ is a countable alphabet, 
$\mathbb{B}$ is the Boolean algebra, 
$\mathbb{S}$ is a commutative semiring, 
$q \xrightarrow{\dots} q'$ is a transition between states $q$ and $q'$.}
\label{fig:hierarchy}  
\end{figure}
%
%This approach is close to the case of 
%Symbolic Automata (SA)~\cite{dAntoniVeanes17CAV,dAntoni21CACM}, 
%except that the domain for weight values is not restricted to be Boolean, 
%like for the guards in the rules of SA, 
%but can be an arbitrary commutative semiring (assuming some restrictions).
%
The models presented here are finite automata called symbolic-weighted (\SWA),
transducers (\SWT), and pushdown automata
with a visibly restriction~\cite{AlurMadhusudan09nested} (\SWVPA).
%\wrt an edit distance
The latter model of automata computes on \emph{nested words}~\cite{AlurMadhusudan09nested}, 
a structured form of words parenthesized with markup symbols, 
corresponding to a linearization of trees.
In the context of parsing, they can represent (weighted) AST of CF grammars.
More precisely, a \SWVPA $A$ associates a weight value $A(t)$ % \in \Semiring$ 
to a given a nested word $t$, which is the linearization of an AST. %representing a parse tree.
%
On the other hand, 
a \SWT can define a distance $T(s, t)$ between finite words $s$ and $t$
over infinite alphabets. % following~\cite{Mohri03EDWA}.
Then, the \emph{SW-parsing} problem aims at %computing 
finding $t$ minimizing 
$T(s, t) \otimes A(t)$ (\wrt the ranking defined by $\oplus$).
The latter value is called the distance between $s$ and $A$ in~\cite{Mohri03EDWA}.
%
Like weighted-parsing 
methods~\cite{Goodman99SemiringParsing,Nederhof03weightedParsing,MorbitzVogler19weighted-parsing}, 
our approach proceeds in two steps, 
based on properties of the \swM. 
The first step is an intersection 
(Bar-Hillel construction~\cite{GruneJacobs08parsing}) 
\florent{chap. intersection in~\cite{GruneJacobs08parsing}} 
where, given a \SWT $T$, a \SWVPA $A$, and an input word $s$, 
a \SWVPA $A_{T, s}$ is built, such that for all $t$, $A_{T, s}(t) = T(s, t) \otimes A(t)$.
In the second step, a best AST $t$ is found by applying to $A_{T, s}$ 
a best search algorithm similar to the shortest distance 
in graphs~\cite{Mohri02semiring,Huang05kbest}.
%In expressiveness, they are equivalent to weighted CFG. 
%and can be used in a general approach for parsing over infinite input alphabets.
%
%Let $A$ be a \SWVPA, associating $A(w) \in \Semiring$ 
%to a given a nested word $w$ (representing a parse tree),
%and let a \SWT compute a distance $d$, in $\Semiring$, 
%between 2 strings over respectively an infinite input alphabet and the 
%same (infinite) alphabet of $A$.
%Then, the problem of Symbolic Weighted Parsing is, 
%given an input string $s$, to find a nested word $w$ minimizing 
%(according to the ranking defined by $\oplus$)
%the distance $d(s, w) \otimes A(w)$ between $s$ and $A$, 
%as defined in~\cite{Mohri03EDWA}.

% First one general edit-distance is defined by a weighted word 
% transducer~\cite{Mohri}
% %Symbolic automata transducers are extended models~\cite{VeanesdAntoniJACM}
% %dealing with infinite set of input symbols...
% value in a semiring...
\florent{expressiveness: VPA have restricted equality test. 
        comparable to pebble automata? $\to$ conclusion}

The main contributions of the paper are: 
($i$)~the introduction of automata, \SWA, transducers, \SWT (Section~\ref{section:SWA}),
and visibly pushdown automata \SWVPA (Section~\ref{sec:SWVPA}),
generalizing the corresponding classes of symbolic and weighted models, 
%for the weighted computation on (nested) words over infinite alphabets;  
($ii$)~a polynomial best-search algorithm for \SWVPA, %(Section~\ref{sec:best})
% a framework for parsing over infinite alphabets, 
and ($iii$)~a uniform framework (Section~\ref{sec:parsing}) for parsing over infinite alphabets, 
the keys to which are 
($iii.a$)~the \SWT-based definition of generic edit distances between input and output (yield) words,
and ($iii.b$)~the use, convenient in this context, 
of nested words, and \SWVPA, 
instead of syntax trees and grammars. %(\S~\ref{sec:trees}).
%
%Finally, Section~\ref{sec:transcription} describes
%the implemented application %of this approach 
%to automated music transcription
%that motivated this study.

>>>>

% !TEX root = main.tex
%\todo
% Introduction
%
Parsing %variant of membership problem for formal languages
%- given a language model and an input word $s$, compute a derivation of the model that yields $s$.
is the problem %process
of structuring a linear representation on input %typically
(a finite word), according to a language model. % (a formal grammar).
% natural language, programming language,
%
Most of the context-free parsing approaches~\cite{GruneJacobs08parsing}
assume a finite and reasonably small input alphabet. %models and algorithms
Such a restriction makes perfect sense in the context of
NLP tasks such as constituency parsing,
or of programming languages compilers or interpreters.
Considering large or infinite alphabets can however be of
practical interest, %in other cases.
%
for instance, when dealing with large characters encodings such as UTF-16, % processing strings in
\eg for vulnerability detection in Web-applications~\cite{dAntoni21CACM},
%
for the analysis (\eg validation or filtering)
of data streams or serialization of structured documents
(with textual or numerical attributes)~\cite{Segoufin06csl},
or for processing timed execution traces~\cite{Bouyer03algebraic}.
% algebraic definition of a class of data languages
% (notion of monoid recognizability, based on registers, comparable to Bojancszik et al. data words)
%

The latter case is related to a study that motivated the present work:
automated music transcription. Most representations of music are
essentially linear. This is true for audio files,
but also for widely used symbolic representations like MIDI.
Such representations ignore the hierarchical structures that frame the
conception of music, at least in the western area. These structures, on the other hand,
are present, either explicitly  or implicitly,
in music notation~\cite{Gould11Notation}: music scores are partitioned in measures, measures
in beats, and so on so forth. Music events do not occur at arbitrary timestamps,
but respect a discrete division of the  timeline incurred by
these recursive divisions.
 The \emph{transcription problem} takes
a linear input (audio or MIDI) and aims at re-constructing these structures
by mapping input events to this hierarchical rhythmic space.
It can therefore be stated as a parsing problem~\cite{foscarin:hal-01988990},
over an infinite alphabet of timed events.

@@

Other extensions of language models  %(automata and grammars)
help in dealing with non-determinism, by the computation of weight values.
With an ambiguous grammar, there may exist several derivations
(\emph{abstract syntax trees} -- AST) % \ie the result of an analyze.
yielding one input word. % structuring a word,
The association of one weight value %in semiring domain
to each AST permits to select a best one (or $n$ bests). % a fixed number of bests
% = ...a ranking of ASTs
This is roughly the principle of \emph{weighted parsing}
approaches~\cite{Goodman99SemiringParsing,Nederhof03weightedParsing,MorbitzVogler19weighted-parsing}.
In \emph{weighted language models},
like \eg probabilistic context-free grammars % (CFG),
and weighted automata (\WA) \cite{Droste09handbook},
a weight value is associated to each transition rule, % production rule
and the rule's weights can be combined with an
associative product operator~$\otimes$ into the weight of an AST.
A second operator~$\oplus$, associative and commutative,
is moreover used to resolve the ambiguity raised by the existence
of several (in general exponentially many) AST
associated to a given input word.
Typically, $\oplus$ will select the best of two weight values.
%Intuitively,~$\oplus$ selects, or ranking, the syntax trees.
The weight domain, equipped with these two operators shall be, at minima,
a \emph{semiring} %$\Semiring$
where $\oplus$ can be extended to infinite sums,
%\cite{Eilenberg74automata}
such as the Viterbi semiring and the tropical min-plus algebra, see Figure~\ref{fig:semirings}.
\philippe{La figure 2 est citée avant la figure 1 mais apparait longtemps après. A corriger.}
%of domain $\mathbb{R}^+ \cup \{ +\infty\}$,
%where $\oplus$ is min and $\otimes$ is plus .
%by ranking
%making the weight domain a semiring.
%Some efficient specialized parsing algorithms~\cite{Huang05kbest} have been proposed in this context
%% models represented as hypergraphs \cite{Huang05kbest}
%in order to compute the $n$ best syntax trees of a given input word without having to enumerate them all.
%Generally based on dynamic programming, these algorithms rely on
%additional algebraic properties of~$\Semiring$.
%-- see \eg~\cite{Huang05kbest} for some NLP applications.
%The extraction of $n$ best list is useful
%the problem: quantitative parsing or symbolic parsing
%of parsing of words over infinite input alphabet.

In this paper, we present a framework for weighted parsing over infinite input alphabets.
\philippe{Tu fais une différence entre model et automata?}
It is based on \emph{symbolic weighted} finite states language models (\swM),
generalizing both~\SA, with functions into an arbitrary semiring instead of Boolean guards,
and~\WA, by handling infinite alphabets, see Figure~\ref{fig:hierarchy}.
%***
%some weighted language models
%model of weighted CFG
%computing on words over infinite input alphabets %sets of terminal symbol
%- but with finite sets of states and transitions rules.
% non-terminals and production rules).
In the transition rules of \swM models, input symbols appear as variables, %(parameters)
and the weight associated to a transition rule is a function of these variables.
%
\begin{figure}
\centering
\begin{tikzpicture}
\node (NFA) at (0,4) {%
  \(
  \begin{array}{c}
  \mathsf{FA} : \Sigma_{\mathsf{fin}}^* \to \mathbb{B}\\
  q \xrightarrow{a} q' \quad  a \in \Sigma_\mathsf{fin}
  \end{array}
  \)
};
%
\node (WA) at (-2.5,2) {%
  \(
  \begin{array}{c}
  \mathsf{WA} : \Sigma_{\mathsf{fin}}^* \to \mathbb{S}\\
  q \xrightarrow{a, w} q' \quad  a \in \Sigma_\mathsf{fin}, w \in \mathbb{S}
  \end{array}
  \)
};
%
\node (SA) at (2.5,2) {%
  \(
  \begin{array}{c}
  \mathsf{SA} : \Sigma_{\mathsf{inf}}^* \to \mathbb{B}\\
  q \xrightarrow{\phi} q' \quad \phi : \Sigma_\mathsf{inf} \to \mathbb{B}
  \end{array}
  \)
};
%
\node (SWA) at (0,0) {%
  \(
  \begin{array}{c}
  \mathsf{SWA} : \Sigma_{\mathsf{inf}}^* \to \mathbb{S}\\
  q \xrightarrow{\phi} q' \quad \phi : \Sigma_\mathsf{inf} \to \mathbb{S}
  \end{array}
  \)
};
\draw[->] (NFA)--(WA);
\draw[->] (NFA)--(SA);
\draw[->] (WA)--(SWA);
\draw[->] (SA)--(SWA);
%\begin{array}{c} \mathsf{NFA} : \Sigma^* \to \mathbb{B} \end{array}
\end{tikzpicture}
\caption{Classes of Symbolic/Weighted Automata.
$\Sigma_\mathsf{fin}$ is a finite alphabet,
$\Sigma_\mathsf{inf}$ is a countable alphabet,
$\mathbb{B}$ is the Boolean algebra,
$\mathbb{S}$ is a commutative semiring,
$q \xrightarrow{\dots} q'$ is a transition between states $q$ and $q'$.}
\label{fig:hierarchy}
\end{figure}
%
%This approach is close to the case of
%Symbolic Automata (SA)~\cite{dAntoniVeanes17CAV,dAntoni21CACM},
%except that the domain for weight values is not restricted to be Boolean,
%like for the guards in the rules of SA,
%but can be an arbitrary commutative semiring (assuming some restrictions).
%
\philippe{Tu veux dire: les modèles formels que tu combines?}
The models presented here are finite automata called symbolic-weighted (\SWA),
transducers (\SWT), and pushdown automata
with a visibly restriction~\cite{AlurMadhusudan09nested} (\SWVPA).
%\wrt an edit distance
The latter model of automata operates on \emph{nested words}~\cite{AlurMadhusudan09nested},
a structured form of words parenthesized with markup symbols,
corresponding to a linearization of trees.
In the context of parsing, they can represent (weighted) AST of CF grammars.
More precisely, a \SWVPA $A$ associates a weight value $A(t)$ % \in \Semiring$
to a given nested word $t$, which is the linearization of an AST. %representing a parse tree.
%
On the other hand,
a \SWT can define a distance $T(s, t)$ between finite words $s$ and $t$
over infinite alphabets. % following~\cite{Mohri03EDWA}.
Then, the \emph{SW-parsing} problem aims at %computing
finding $t$ minimizing
$T(s, t) \otimes A(t)$ (\wrt the ranking defined by $\oplus$), given an input word $s$.
The latter value is called the distance between $s$ and $A$ in~\cite{Mohri03EDWA}.
%

Like weighted-parsing
methods~\cite{Goodman99SemiringParsing,Nederhof03weightedParsing,MorbitzVogler19weighted-parsing},
our approach proceeds in two steps,
based on properties of the \swM.
The first step is an intersection
(Bar-Hillel construction~\cite{GruneJacobs08parsing})
\florent{chap. intersection in~\cite{GruneJacobs08parsing}}
where, given a \SWT $T$, a \SWVPA $A$, and an input word $s$,
a \SWVPA $A_{T, s}$ is built, such that for all $t$, $A_{T, s}(t) = T(s, t) \otimes A(t)$.
\philippe{The notation  $A_{T, s}$ has  not been introduced so far. It is not clear why $T$ is a parameter there}
In the second step, a best AST $t$ is found by applying to $A_{T, s}$
a best search algorithm similar to the shortest distance
in graphs~\cite{Mohri02semiring,Huang05kbest}.
%In expressiveness, they are equivalent to weighted CFG.
%and can be used in a general approach for parsing over infinite input alphabets.
%
%Let $A$ be a \SWVPA, associating $A(w) \in \Semiring$
%to a given a nested word $w$ (representing a parse tree),
%and let a \SWT compute a distance $d$, in $\Semiring$,
%between 2 strings over respectively an infinite input alphabet and the
%same (infinite) alphabet of $A$.
%Then, the problem of Symbolic Weighted Parsing is,
%given an input string $s$, to find a nested word $w$ minimizing
%(according to the ranking defined by $\oplus$)
%the distance $d(s, w) \otimes A(w)$ between $s$ and $A$,
%as defined in~\cite{Mohri03EDWA}.

% First one general edit-distance is defined by a weighted word
% transducer~\cite{Mohri}
% %Symbolic automata transducers are extended models~\cite{VeanesdAntoniJACM}
% %dealing with infinite set of input symbols...
% value in a semiring...
\florent{expressiveness: VPA have restricted equality test.
        comparable to pebble automata? $\to$ conclusion}
The main contributions of the paper are:
($i$)~the introduction of automata, \SWA, transducers, \SWT (Section~\ref{section:SWA}),
and visibly pushdown automata \SWVPA (Section~\ref{sec:SWVPA}),
generalizing the corresponding classes of symbolic and weighted models,
%for the weighted computation on (nested) words over infinite alphabets;
($ii$)~a polynomial best-search algorithm for \SWVPA, %(Section~\ref{sec:best})
% a framework for parsing over infinite alphabets,
and ($iii$)~a uniform framework (Section~\ref{sec:parsing}) for parsing over infinite alphabets,
the keys to which are
($iii.a$)~the \SWT-based definition of generic edit distances between input and output (yield) words,
and ($iii.b$)~the use, convenient in this context,
of nested words, and \SWVPA,
instead of syntax trees and grammars. %(\S~\ref{sec:trees}).
%
%Finally, Section~\ref{sec:transcription} describes
%the implemented application %of this approach
%to automated music transcription
%that motivated this study.
